---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
library("ggplot2")
temp<-tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", temp)

```

## What is mean total number of steps taken per day?

```{r}
activities<- read.csv(unz(temp,"activity.csv"))
totalStepsPerDay<-aggregate(activities$steps, by=list(activities$date),FUN=sum)
totalStepsPerDay$date<-strptime(totalStepsPerDay$Group.1,"%Y-%m-%d")
mx<-mean(totalStepsPerDay$x,na.rm=T)
qplot( x, data=totalStepsPerDay, xlab="Steps", na.rm=T, bins=50)+geom_histogram(na.rm=T,bins=50)+geom_vline(aes(xintercept=mx),   color="red", linetype="dashed", size=1)+geom_text(aes(x=mx, label=paste("\nMeans=",round(mx,digits=0)), y=4), colour="red", angle=90)

```

Mean of total number of steps taken per day = `r toString(round(mx,digits=0))` steps.

Median of total number of steps taken per day = `r toString(median(totalStepsPerDay$x, na.rm=T))` steps.



## What is the average daily activity pattern?
```{r}

totalStepsPerIntervalBlock<-aggregate(activities$steps, by=list(activities$interval), FUN=sum, na.rm=T, na.action=NULL)
colnames(totalStepsPerIntervalBlock)<-c("Intervals", "Steps")
mx<-totalStepsPerIntervalBlock[which.max(totalStepsPerIntervalBlock$Steps),]

ggplot(totalStepsPerIntervalBlock, aes(Intervals, Steps))+geom_line()+geom_vline(aes(xintercept=mx$Intervals),   color="red", linetype="dashed", size=1)+geom_text(aes(x=mx$Intervals, label=paste("\nMax at interval ",mx$Intervals," = ",mx$Steps, " steps" ), y=3000), colour="red", angle=90)

```

Interval `r toString(mx$Intervals)` contains the maximum number of steps of `r toString(mx$Steps)`

## Imputing missing values

Total number of missing values in dataset = `r toString(sum(is.na(activities$steps)))` rows

Preparing a list of candidate values to replace missing values in data set using the mean steps per interval block;

```{r}
meanStepsPerIntervalBlock<-aggregate(activities$steps,by=list(activities$interval), FUN=mean,na.rm=T, na.action=NULL)
colnames(meanStepsPerIntervalBlock)<-c("Intervals", "Steps")
meanStepsPerIntervalBlock$roundMeanSteps<-round(meanStepsPerIntervalBlock$Steps, digits=0)
```


Replacing the missing values with the candidate values that have been prepared earlier;

```{r}
cleanUpActivities<-activities
for(i in 1:nrow(cleanUpActivities)){
  if(is.na(cleanUpActivities[i,"steps"])){
    cleanUpActivities[i,"steps"]<-meanStepsPerIntervalBlock[meanStepsPerIntervalBlock$Intervals==cleanUpActivities[i, "interval"],"roundMeanSteps"]
  }
}
```

Plotting the histogram with the cleaned up dataset;

```{r}
newTotalStepsPerDay<-aggregate(cleanUpActivities$steps, by=list(cleanUpActivities$date),FUN=sum)
newTotalStepsPerDay$date<-strptime(newTotalStepsPerDay$Group.1,"%Y-%m-%d")
mx<-mean(newTotalStepsPerDay$x,na.rm=T)
qplot( x, data=newTotalStepsPerDay, xlab="Steps", na.rm=T, bins=50)+geom_histogram(na.rm=T,bins=50)+geom_vline(aes(xintercept=mx),   color="red", linetype="dashed", size=1)+geom_text(aes(x=mx, label=paste("\nMeans=",round(mx,digits=0)), y=4), colour="red", angle=90)

```

Mean of total number of steps taken per day in the new, cleaned up dataset = `r toString(round(mx,digits=0))` steps.

Median of total number of steps taken per day in the new, cleaned up dataset= `r toString(median(newTotalStepsPerDay$x, na.rm=T))` steps.

While there is no significant impact of imputing missing data on the mean but the median value has reduced by `r toString(median(totalStepsPerDay$x, na.rm=T)-median(newTotalStepsPerDay$x, na.rm=T))`.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
activities$weekstate<-ifelse(weekdays(as.Date(activities$date),abbr=T)!="Sat"&weekdays(as.Date(activities$date),abbr=T)!="Sun","weekday","weekend")


newTotalStepsPerIntervalBlock<-aggregate(activities$steps, by=list(activities$interval, activities$weekstate), FUN=sum, na.rm=T, na.action=NULL)

colnames(newTotalStepsPerIntervalBlock)<-c("interval","weekstate", "steps")


ggplot(newTotalStepsPerIntervalBlock,aes(interval, steps))+geom_line(color="blue")+facet_wrap(~weekstate, ncol=1)+labs(y="Number of steps", x="Interval")
#+theme(strip.text=element_text(angle=180))
```

Based on the charts, there is signifcant drop of activity volume on weekends as compared to weekdays, especially during the usual rush hours in the morning between 6 am - 9 am and end of the day between 6 pm - 7 pm.